opXRD: Open Experimental Powder X-ray Diffraction Database

### 2024-11-05 chapters/01_introduction.tex
Added paragraphs in introduction

\begin{figure*}[!ht]
    \centering
    \includegraphics{figures/overview.pdf}
    \caption{Overview of the data collection pipeline. Datasets are submitted using an online submission form or our submission helper software. After post-processing and data homogenization, we create a Zenodo entry for each user submission and subsequently include the submission in the opXRD database.}
    \label{fig:overview}
\end{figure*}

% High throughput experimentation
The advent of high-throughput experiments holds the prospect of significantly accelerating the speed of materials discovery. \cite{Liu2019}
The synthesis and characterization of novel materials is becoming increasingly efficient and automated,
increasing the throughput of samples in experimentation pipelines. \cite{MacLeod2019, Ludwig2019, Ozaki2020}

% On Rietveld refinement
After fabricating a new material, a number of analysis techniques can be used to characterize the sample.
The most important method to determine the crystal structure of a new material is powder X-ray diffraction. 
To analyze pXRD data, an initial guess is optimized using Rietveld refinement. In this process, an initial crystal structure including unit cell parameters is provided by the operator of the analysis software and fitted to the observed diffractogram in an iterative optimization process based on simulated forward prediction of expected pXRD diffractograms given the current crystal structure.\cite{Dinnebier2019, Cano2021}
As Rietveld refinement is a local optimization method, the result of the refinement procedure is generally only as good as the initial guess it was provided with.
%Therefore, it strongly relies on the experience of the operator to provide an accurate initial guess of the crystal structure.

% Manual vs (ML) automated Refinement
Employing manually operated Rietveld refinement, including determining a good initial structure for the optimization, is not without issues.
It is certainly not scalable to the degree required to keep up with advances in throughput and efficiency in other steps of the experimentation pipeline.
Automation of the Rietveld refinement requires the automation of finding initial guess structures, which is highly non-trivial, and the automation of the iterative optimization procedure, which can be done using black-box optimization algorithms \cite{Ozaki2020}. 
Initial guesses for Rietveld refinement are often obtained by matching the observed diffractogram to patterns from a database. Commonly used databases include the Powder Diffraction File\cite{GatesRector2019}, the Inorganic Crystal Structure Database \cite{Belsky2002}, and the Crystallography Open Database \cite{Grazulis2009}.
An initial guess obtained from such a database is by no means guaranteed to lead to an accurate structure solution, even if it is iteratively revised, either manually or using blackbox optimization.\cite{Ozaki2020}
Especially for novel observed structure types, simply matching to a database of known structures will not yield good results. Therefore, more complex crystal structure solution approaches need to be applied to find suitable crystal structure guesses.
Machine learning has the potential to speed up the manual analysis of powder diffractograms and keep pace with an automated high-throughput experimentation environment. \cite{Agrawal2019, Surdu2023}
Models can be trained to predict crystal structure information directly given a diffractogram, or models can be used to automate the conventional workflow of finding an initial guess \cite{Surdu2023} and running ML-automated Rietveld refinement \cite{Feng2019}.
So far, however, due to an absence of labeled datasets with experimental diffractograms\cite{Wang2020}, machine learning in this domain has largely relied on diffractograms simulated from known structures\cite{Park2017, Lee2023} or, most recently, from generated synthetic crystals \cite{Schopmans2023}

@@

At the moment, the number of openly available datasets for experimental powder data is very limited.
The main sources for experimental patterns are XXX, YYY, and ZZZ.
Further commercial datasets such as KKK and LLL exist, but license fees and restrictions in their use to train publishable models or even publish the training data limit their usability.
Collecting a labeled dataset of raw experimental powder X-ray diffraction data of sufficient size to directly train a model that predicts crystal structure information is likely out of reach.
The most advanced ML models currently are trained on approximately $10^7 - 10^8$ simulated diffractograms, which is - to the best of our knowledge - three orders of magnitude larger than the largest currently curated experimental dataset and even x orders of magnitude larger than the number of unlabeled diffractograms in the opXRD dataset we present here.

However, even small labeled and also unlabeled experimental diffractograms hold significant value in practice. First of all, labeled data can be used to better test and benchmark existing and new automated analysis approaches, which yields a better comparison than when using the very limited currently available datasets, such as the commonly used RRUFF mineral dataset. \cite{Armbruster2015}

Furthermore, even completely unlabeled data can enable machine learning researchers to evaluate how closely their simulations represent experimental data, identify what features are missing, and modify their simulation algorithms accordingly to arrive at a simulation that mirrors how data appears in reality.
One can bring the simulated patterns closer to experiment by improving the simulation itself by properly modeling more of the observed physical effects such as preferred orientation, peak shapes resulting from different sources, etc.
Additionally, some of the observed imperfections, such as background and noise, can be modeled using machine learning approaches.
This is similar to the concept of style transfer\cite{Gatys2016, Ganin2015} found in the domain of machine learning applied to images.
Here, the objective is to obtain a classification or regression model when only labeled images of one style (simulated data) and unlabeled or partially labeled images of another style (experimental data) are available.

In this publication, we provide a new open powder X-ray diffraction (opXRD) dataset that collects a broad range of patterns from experiment, exceeding the currently available amount of openly accessible experimental patterns by XX.
Our opXRD dataset has been curated by collecting the accumulated powder data from multiple large research groups and institutions with high-throughput XRD facilities. It contains pXRD patterns from single and multiphase materials from a wide variety of materials classes, including XXX, YYY, and ZZZ. Details about the experiments of contributing research groups are discussed in this paper, along with a detailed description of how to contribute further data and how to use the data.

>>>>

\begin{figure*}[!htb]
    \centering
    \includegraphics{figures/overview.pdf}
    \caption{Overview of the data collection pipeline. Datasets are submitted using an online submission form or our submission helper software. After post-processing and data homogenization, we create a Zenodo entry for each user submission and subsequently include the submission in the opXRD database.}
    \label{fig:overview}
\end{figure*}

% High throughput experimentation
The advent of high-throughput experiments holds the prospect of significantly accelerating the speed of materials discovery. \cite{Liu2019} The synthesis and characterization of novel materials are becoming increasingly efficient and automated, increasing the throughput of samples in experimentation pipelines. \cite{MacLeod2019, Ludwig2019, Ozaki2020} \\

% On Rietveld refinement
After fabricating a new material, a number of analysis techniques can be used to characterize the sample. One method that can be used to determine the crystal structure of a new material is powder X-ray diffraction. To determine crystal structure from pXRD data, an initial guess is optimized using Rietveld refinement. In this process, an initial crystal structure including unit cell parameters is provided by the operator of the analysis software and fitted to the observed diffractogram in an iterative optimization process based on simulated forward prediction of expected pXRD diffractograms given the current crystal structure. \cite{Dinnebier2019, Cano2021} As Rietveld refinement is a local optimization method, the result of the refinement procedure is only as good as the initial guess it was provided with. \\

% Manual vs (ML) automated Refinement
Employing manually operated Rietveld refinement, including determining a good initial structure for the optimization, is not without issues.
It is certainly not scalable to the degree required to keep up with advances in throughput and efficiency in other steps of the experimentation pipeline.
Automation of the Rietveld refinement requires the automation of finding initial guess structures, which is highly non-trivial, and the automation of the iterative optimization procedure, which can be done using black-box optimization algorithms \cite{Ozaki2020}. 
Initial guesses for Rietveld refinement are often obtained by matching the observed diffractogram to patterns from a database. Commonly used databases include the Powder Diffraction File\cite{GatesRector2019}, the Inorganic Crystal Structure Database \cite{Belsky2002}, and the Crystallography Open Database \cite{Grazulis2009}. An initial guess obtained from such a database is by no means guaranteed to lead to an accurate structure solution, even if it is iteratively revised, either manually or using blackbox optimization.\cite{Ozaki2020} Especially for novel observed structure types, simply matching to a database of known structures will not yield good results. Therefore, more complex crystal structure solution approaches need to be applied to find suitable crystal structure guesses. \\

Machine learning has the potential to speed up the manual analysis of powder diffractograms and keep pace with an automated high-throughput experimentation environment. \cite{Agrawal2019, Surdu2023}
Models can be trained to predict crystal structure information directly given a diffractogram, or models can be used to automate the conventional workflow of finding an initial guess \cite{Surdu2023} and running ML-automated Rietveld refinement \cite{Feng2019}.
So far, however, due to an absence of labeled datasets with experimental diffractograms\cite{Wang2020}, machine learning in this domain has largely relied on diffractograms simulated from known structures\cite{Park2017, Lee2023} or, most recently, from generated synthetic crystals. \cite{Schopmans2023} \\

@@

At the moment, the number of openly available datasets for experimental powder data is very limited. The main sources for experimental patterns are XXX, YYY, and ZZZ. Further commercial datasets such as KKK and LLL exist, but license fees and restrictions in their use to train publishable models or even publish the training data limit their usability. Collecting a labeled dataset of raw experimental powder X-ray diffraction data of sufficient size to directly train a model that predicts crystal structure information is likely out of reach. The most advanced ML models currently are trained on approximately $10^7 - 10^8$ simulated diffractograms, which is - to the best of our knowledge - three orders of magnitude larger than the largest currently curated experimental dataset and even x orders of magnitude larger than the number of unlabeled diffractograms in the opXRD dataset we present here. \\

However, even small labeled and also unlabeled experimental diffractograms hold significant value in practice. First of all, labeled data can be used to better test and benchmark existing and new automated analysis approaches, which yields a better comparison than when using the very limited currently available datasets, such as the commonly used RRUFF mineral dataset. \cite{Armbruster2015} \\

Furthermore, even completely unlabeled data can enable machine learning researchers to evaluate how closely their simulations represent experimental data, identify what features are missing, and modify their simulation algorithms accordingly to arrive at a simulation that mirrors how data appears in reality.
One can bring the simulated patterns closer to experiment by improving the simulation itself by properly modeling more of the observed physical effects such as preferred orientation, peak shapes resulting from different sources, etc.
Additionally, some of the observed imperfections, such as background and noise, can be modeled using machine learning approaches. This is similar to the concept of style transfer\cite{Gatys2016, Ganin2015} found in the domain of machine learning applied to images. Here, the objective is to obtain a classification or regression model when only labeled images of one style (simulated data) and unlabeled or partially labeled images of another style (experimental data) are available. \\

In this publication, we provide a new open powder X-ray diffraction (opXRD) dataset that collects a broad range of patterns from experiment, exceeding the currently available amount of openly accessible experimental patterns by XX.
Our opXRD dataset has been curated by collecting the accumulated powder data from multiple large research groups and institutions with high-throughput XRD facilities. It contains pXRD patterns from single and multiphase materials from a wide variety of materials classes, including XXX, YYY, and ZZZ. Details about the experiments of contributing research groups are discussed in this paper, along with a detailed description of how to contribute further data and how to use the data. \\

### 2024-11-05 chapters/02_existing_datasets.tex
Commited external author contribution
Added listing of all found Crystallography databases
Test commit
Commited INT contribution paragraph
Added paragraphs in introduction

% Crystal structures with corresponding simulated diffractograms;
% ->  ICSD (https://icsd.products.fiz-karlsruhe.de/) 
% - 230k structures;
% - Proprietary

% -> PDF (https://www.icdd.com/) -> 
%- 1200k "material data sets" (structures?) 
% + 19 k entries with "Experimental raw data digital patterns"; 
%  Proprietary (% PDF pricing: https://www.icdd.com/assets/files/2023-Purchase-Options.pdf)
%The PDF itself lists the following data sources:

\paragraph{Existing datasets} TODO: Summarize the existing datasets shortly, including number of entries, etc.

\begin{itemize}
\item RRUFF
\item Powder Diffraction File (ICDD)
%\begin{itemize}
%    \item ICDD Powders (00) – International Centre for Diffraction Data Newtown Square, PA, USA
%    \item ICSD (01) – FIZ Karlsruhe – Leibniz Institute for Information Infrastructure (FIZ), Karlsruhe, Germany – Inorganic Crystal Structure Database
%    \item NIST (03) – National Institute of Standards and Technology, Gaithersburg, Maryland, USA
%    \item MPDS (04) – Material Phases Data System, Vitznau, Switzerland – Linus Pauling File (LPF)
%    \item ICDD Single Crystal Data (05) – International Centre for Diffraction Data Newtown Square, PA, USA
%    \item TODO: Include how many of those are actually experimental pXRD! If we cannot find this information, we need to ask them directly.
%\end{itemize}
\item Scattered experimental datasets on Zenodo (TODO: roughly estimate size)
\item ...
\end{itemize}

@@

>>>>

% Crystal structures with corresponding simulated diffractograms;
% ->  ICSD 
% - 230k structures;
% - Proprietary

% -> PDF  -> 
%- 1200k "material data sets" (structures?) 
% + 19 k entries with "Experimental raw data digital patterns"; 
%  Proprietary (% PDF pricing: https://www.icdd.com/assets/files/2023-Purchase-Options.pdf)
%The PDF itself lists the following data sources:

@@

% Datset listing (Done on: 20.06.24)
% I would list here *all* crystallography databases but evaluate/determine that they lack or don't provide convenient access to raw powder diffraction data

% Some other Lists of Crystallographic databases
% - IUCR: "Crystallographyic databases and related resources" (https://www.iucr.org/resources/data/databases)
% - Bruno et al. (2017): "Crystallography and Databases" (https://datascience.codata.org/articles/10.5334/dsj-2017-038)
% - University of Virginia "Crystallographic data" (https://guides.lib.virginia.edu/c.php?g=514782&p=3568808)

%->1 PDF5+ (https://www.icdd.com/,https://www.icdd.com/pdf-5/)
% - Data sources of the PDF
% (1.1): ICDD Powders
% (1.2): FIZ ICSD (https://icsd.products.fiz-karlsruhe.de/)
% (1.3): NIST ICSD https://icsd.nist.gov/
% (1.4): Linus Pauling File (https://paulingfile.com/index.php?p=about%20us)
% (1.5): ICDD Single Crystal data
%-> 2 COD (https://www.crystallography.net/cod/)
%-> 3 RRUFF (American Mineralogist Crystal Structure Database)
%-> 4 CCDC (https://www.ccdc.cam.ac.uk/)
%-> 5 The Material Project (https://legacy.materialsproject.org/)
%-> 6 Crystal Lattice Structures (https://www.atomic-scale-physics.de/lattice/)
%-> 7 Crystallographic and Crystallochemical Database for Minerals and their Structural Analogues (https://database.iem.ac.ru/mincryst/)
%-> 8 Mineralogy Database (https://webmineral.com/)
%-> 9 IUCr Raw data letters (https://iucrdata.iucr.org/)
%-> 10 Bilbao Crystallographic server  (https://www.cryst.ehu.es/)
%-> 11 PowBase (http://www.cristal.org/powbase/index.html)
%-> 12 Athena (https://athena.unige.ch/athena/mineral/mineral.html)
%-> 13 Protein database (https://www.rcsb.org/)
%-> 14 SDPD (Structure Determination from Powder Diffraction - Database)
%-> 15 Crystal Met (https://cds.dl.ac.uk/cds/datasets/crys/mdf/llmdf.html)
%-> 16 NAKB (https://www.nakb.org/)
%-> 17 Pearson's Crystal data (https://www.crystalimpact.com/pcd/)
%-> 18 Biological macromolecule Crystallization Database (http://bmcd.ibbr.umd.edu/)

%TODO: Also include plans for NOMAD and highlight their efforts

%TODO: Summarize the existing datasets shortly, including number of entries, etc.

\paragraph{Existing powder diffraction datasets} To contextualize opXRD within the current environment of experimental powder diffraction data, the following list provides an overview of the largest crystal structure databases that include powder diffraction data.

\begin{itemize}
\item \textbf{Powder Diffraction File:} \footnote{More information on the Powder Diffraction File can be found under \url{https://www.icdd.com/pdf-5/}.} The Powder Diffraction File (PDF), published and maintained by the International Center for diffraction data, is currently the largest collection of powder diffraction data available to researchers. Though its latest release contains over a million entries, almost all are simulated patterns. Only 19000 entries represent powder diffraction data originating from experiments, and of those, only about 11000 come with accompanying atomic coordinates. It is widely used as a reference database in Rietveld refinement to identify crystal structures from which the refinement process can begin based on their powder diffraction patterns. As of November 2024, the latest release of the PDF, the PDF5+, is available to researchers through a purchase starting at a price point of \$6,265.00.

\item \textbf{Crystallography Open Database:}\footnote{More information on the crystallography open database can be found under \url{https://www.crystallography.net/cod/}.} The Crystallography Open Database (COD) is an open-access collection of crystal structures. It provides over 300.000 crystal structures in the form of .cif files. Of these files, 1063 contained the experimental powder diffraction data that was used to determine the underlying crystal structures of the investigated samples.

\item Scattered experimental datasets on Zenodo (TODO: roughly estimate size)
\item RRUFF
\item ...
\end{itemize}

Some major crystal structure databases, such as the ICSD, do not appear in the above list, as they do not include powder diffraction data. This is because most structure solutions are achieved through single-crystal diffraction rather than powder diffraction.

### 2024-11-05 main.tex
Added paragraphs in introduction

\affil[1]{Institute of Theoretical Informatics, Karlsruhe Institute of Technology, 76131 Karlsruhe, Germany}
\affil[2]{Institute of Nanotechnology, Karlsruhe Institute of Technology, 76131 Karlsruhe, Germany}
\affil[3]{Department of Chemical Engineering and Materials Science, University of Southern California, Los Angeles CA 90089, USA}
\affil[4]{Lawrence Berkeley National Laboratory, Molecular Foundry Division, Berkeley 94720 CA, USA}
\affil[5]{Empa-Swiss Federal Laboratories
for Materials Science and Technology, CH-8600 Dübendorf,
Switzerland}
\affil[6]{Engler-Bunte-Institut \& Institute of Catalysis
Research and Technology, Karlsruhe Institute of Technology, Karlsruhe, Germany}
\affil[7]{Chimie ParisTech, PSL University, CNRS, Institut de Recherche de Chimie Paris, 75005 Paris, France}
\affil[8]{Guangzhou Municipal Key Laboratory of Materials Informatics,  Hong Kong University of
Science and Technology (Guangzhou), Guangzhou 511400, Guangdong, China}

>>>>

\affil[1]{Institute of Theoretical Informatics, Karlsruhe Institute of Technology, 76131 Karlsruhe, Germany}
\affil[2]{Institute of Nanotechnology, Karlsruhe Institute of Technology, 76131 Karlsruhe, Germany}
\affil[3]{Department of Chemical Engineering and Materials Science, University of Southern California, Los Angeles CA 90089, USA}
\affil[4]{Lawrence Berkeley National Laboratory, Molecular Foundry Division, Berkeley 94720 CA, USA}
\affil[5]{Empa-Swiss Federal Laboratories
for Materials Science and Technology, CH-8600 Dübendorf,
Switzerland}
\affil[6]{Engler-Bunte-Institut \& Institute of Catalysis
Research and Technology, Karlsruhe Institute of Technology, Karlsruhe, Germany}
\affil[7]{Chimie ParisTech, PSL University, CNRS, Institut de Recherche de Chimie Paris, 75005 Paris, France}
\affil[8]{Guangzhou Municipal Key Laboratory of Materials Informatics,  Hong Kong University of
Science and Technology (Guangzhou), Guangzhou 511400 Guangdong, China}
